---
title: "Forecasting age-stratified norovirus gastroenteritis counts using `surveillance::hhh4`"
author: "Sebastian Meyer"
date: "`r Sys.Date()`"
output:
    rmarkdown::html_vignette:
        fig_width: 6
        fig_height: 4
        toc: TRUE
vignette: >
  %\VignetteIndexEntry{Forecasting age-stratified norovirus gastroenteritis counts using 'surveillance::hhh4'}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{ggplot2, surveillance, hhh4contacts, fanplot}
---

```{r setup_knitr, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, error = FALSE,
                      fig.align = "center", dev.args = list(pointsize = 10))
```

```{r setup}
library("HIDDA.forecasting")
library("ggplot2")
```

In this vignette, we use modelling and forecasting methods provided by:
```{r}
library("surveillance")
```

The corresponding software reference is:
```{r, echo = FALSE, results = "asis"}
cat("<blockquote>")
print(citation(package = "surveillance", auto = TRUE), style = "html")
cat("</blockquote>\n")
```


## Data

We use age-stratified norovirus surveillance data from Berlin, Germany, as
provided by:

```{r}
library("hhh4contacts")
```

These data (as well as spatially disaggregated versions) have been originally analyzed in:

> Meyer S and Held L (2017): "Incorporating social contact data in
> spatio-temporal models for infectious disease spread".
> *Biostatistics*, **18**(2), pp. 338--351.
> DOI: [10.1093/biostatistics/kxw051](https://doi.org/10.1093/biostatistics/kxw051)


```{r BNV}
BNV <- noroBE(by = "agegroup", agegroups = c(1, 2, 2, 4, 4, 2),
              timeRange = c("2011-w27", "2016-w26"))
BNV
```

Using the default plot method for "sts" objects:

```{r BNV_stsplot}
plot(BNV)
```

Using the `autoplot` variant to plot the weekly incidence (**ggplot2**
version), also adding some extra markers to the plot:

```{r BNV_ggplot}
autoplot(BNV, population = 100000/sum(pop2011)) +
    geom_bar(aes(fill = epochInYear %in% c(52, 1)),
             stat = "identity", show.legend = FALSE) +
    scale_fill_manual(values = c("black", "red")) +
    geom_vline(aes(xintercept = as.numeric(date)[4*52] + .5), linetype = 2) +
    ylab("Incidence [per 100 000 inhabitants]")
```

We will use the first four seasons, from week 2011/27 to week 2015/26, as
training data, and compute and assess forecasts during the following year
(2015/27 to 2016/26).

These weekly counts of reported cases are stratified by six age groups:

```{r}
(NGROUPS <- ncol(BNV))
(GROUPS <- colnames(BNV))
```

The "neighbourhood" slot of the dataset contains a social contact matrix
estimated from the German subset of the
[POLYMOD survey](http://cordis.europa.eu/project/rcn/79211_en.html):

```{r}
neighbourhood(BNV)
```

Each entry gives the mean number of contact persons of a certain age
group a participant (of a certain age group) reports on a randomly
assigned day.

R package **socialmixr** ...


## Modelling



## One-week-ahead forecasts


## Long-term forecasts

