---
title: "Forecasting Swiss ILI surveillance counts using `surveillance::hhh4`"
author: "Sebastian Meyer"
date: "`r Sys.Date()`"
output:
    rmarkdown::html_vignette:
        fig_width: 6
        fig_height: 4
        toc: TRUE
vignette: >
  %\VignetteIndexEntry{Forecasting Swiss ILI surveillance counts using 'surveillance::hhh4'}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{ggplot2, surveillance, fanplot}
---

```{r setup_knitr, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, error = FALSE,
                      fig.align = "center", dev.args = list(pointsize = 10))
```

```{r setup}
library("HIDDA.forecasting")
library("ggplot2")
source("setup.R")  # define test periods (OWA, TEST)
```


## Modelling

```{r}
library("surveillance")
CHILI.sts <- sts(observed = CHILI,
                 epoch = as.integer(index(CHILI)), epochAsDate = TRUE)
```

```{r hhh4fit}
(weeksInYear <- table(year(CHILI.sts)))
mean(weeksInYear)  # long-term average is 52.1775
f1 <- addSeason2formula(~ 1, period = 52.1775, timevar = "index")
## equivalent: f1 <- addSeason2formula(~ 1, period = 365.2425, timevar = "t")
hhh4fit <- hhh4(stsObj = CHILI.sts,
     control = list(
         ar = list(f = update(f1, ~. + christmas)),
         end = list(f = f1),
         family = "NegBin1",
         data = list(index = 1:nrow(CHILI.sts),
                     christmas = as.integer(epochInYear(CHILI.sts) %in% 52))
     ))
summary(hhh4fit, maxEV = TRUE)
```

Alternatively, use a yearly varying frequency of 52 or 53 weeks for
the sinusoidal time effects:

```{r hhh4fit_varfreq}
f1_varfreq <- ~ 1 + sin(2*pi*epochInYear/weeksInYear) + cos(2*pi*epochInYear/weeksInYear)
hhh4fit_varfreq <- update(hhh4fit,
    ar = list(f = update(f1_varfreq, ~. + christmas)),
    end = list(f = f1_varfreq),
    data = list(epochInYear = epochInYear(CHILI.sts),
                weeksInYear = rep(weeksInYear, weeksInYear)))
AIC(hhh4fit, hhh4fit_varfreq)
```
```{r hhh4fit_maxEV, include = FALSE}
plot(hhh4fit, hhh4fit_varfreq, type = "maxEV",
     matplot.args = list(ylab = expression(lambda[t]), xlim =c(2002,2007)))
```

```{r hhh4fitted_components}
plot(hhh4fit, pch = 20, names = "", ylab = "")
```

```{r hhh4fitted_lambdat, fig.width = 3, fig.height = 2}
qplot(x = 1:53, y = drop(predict(hhh4fit, newSubset = 1:53, type = "ar.exppred")),
      geom = "line", ylim = c(0,1.3), xlab = "week", ylab = expression(lambda[t])) +
    geom_hline(yintercept = 1, lty = 2)
```

```{r}
CHILIdat <- fortify(CHILI)
## add fitted mean and CI to CHILIdat
CHILIdat$hhh4upper <- CHILIdat$hhh4lower <- CHILIdat$hhh4fitted <- NA_real_
CHILIdat[hhh4fit$control$subset,"hhh4fitted"] <- fitted(hhh4fit)
CHILIdat[hhh4fit$control$subset,c("hhh4lower","hhh4upper")] <-
    sapply(c(0.025, 0.975), function (p)
        qnbinom(p, mu = fitted(hhh4fit),
                size = exp(hhh4fit$coefficients[["-log(overdisp)"]])))
```

```{r hhh4fitted}
ggplot(CHILIdat, aes(x=Index, ymin=hhh4lower, y=hhh4fitted, ymax=hhh4upper)) +
    geom_ribbon(fill="orange") + geom_line(col="darkred") +
    geom_point(aes(y=CHILI), pch=20) +
    scale_y_sqrt(expand = c(0,0), limits = c(0,NA))
```


## One-week-ahead forecasts

```{r hhh4owa, eval = !file.exists("hhh4owa.RData"), results = "hide"}
hhh4owa <- oneStepAhead(hhh4fit, range(OWA), type = "rolling", verbose = FALSE)
save(hhh4owa, file = "hhh4owa.RData")
```
```{r, include = FALSE}
load("hhh4owa.RData")
```

```{r hhh4owa_pit, fig.width = 3, fig.height = 3, echo = -1}
par(mar = c(5,5,1,1))
pit(hhh4owa)
```

```{r hhh4owa_caltest, cache = TRUE}
calibrationTest(hhh4owa, which = "dss")
calibrationTest(hhh4owa, which = "logs")
```

```{r hhh4owa_scores}
hhh4owa_scores <- scores(hhh4owa, which = c("dss", "logs"))
summary(hhh4owa_scores)
```

```{r hhh4owa_plot, echo = -1}
par(mar = c(5,5,1,1))
osaplot(
    quantiles = quantile(hhh4owa, probs=1:99/100), probs = 1:99/100,
    observed = hhh4owa$observed, scores = hhh4owa_scores,
    start = OWA[1]+1, xlab = "Week", ylim = c(0,60000),
    fan.args = list(ln = c(0.1,0.9), rlab = NULL)
)
```


## Long-term forecasts

### Example for the first test period

```{r hhh4sim1}
TEST1 <- TEST[[1]]
fit1 <- update(hhh4fit, subset.upper = TEST1[1]-1)
hhh4sim1 <- simulate(fit1, nsim = 1000, seed = 726, subset = TEST1,
                     y.start = observed(CHILI.sts)[TEST1[1]-1,])
```

We can use the plot method provided by **surveillance**:

```{r hhh4sim1_plots, R.options = list(scipen=1)}
par(mfrow=c(1,2))
plot(hhh4sim1, "fan", ylim = c(0,60000), xlab = "", ylab = "",
     epochsAsDate = TRUE, xaxis.tickFreq = list("%m"=atChange, "%Y"=atChange),
     xaxis.labelFreq = list("%Y"=atMedian), xaxis.labelFormat = "%Y")
plot(hhh4sim1, "size", horizontal = FALSE, main = "size of the epidemic",
     ylab = "", observed = list(labels = NULL))
```

There is also an associated `scores` method:

```{r}
summary(scores(hhh4sim1, which = c("dss", "logs")))
```

The log-score is problematic for simulation-based count forecasts.


### For all three test periods

```{r hhh4sims}
hhh4sims <- lapply(TEST, function (testperiod) {
    t0 <- testperiod[1] - 1
    fit0 <- update(hhh4fit, subset.upper = t0)
    sims <- simulate(fit0, nsim = 1000, seed = t0, subset = testperiod,
                     y.start = observed(CHILI.sts)[t0,])
    list(t0 = t0,
         observed = observed(CHILI.sts)[testperiod],
         sims = sims)
})
```

```{r hhh4sims_plot, echo = -1, fig.show = "hold"}
par(mar = c(5,5,1,1))
t(sapply(hhh4sims, function (x) {
    quantiles <- t(apply(x$sims, 1, quantile, probs = 1:99/100))
    scores <- scores(x$sims, which = c("dss", "logs"))
    osaplot(quantiles = quantiles, probs = 1:99/100,
            observed = x$observed, scores = scores[,"dss",drop=FALSE],
            start = x$t0, xlab = "Week", ylim = c(0,60000),
            fan.args = list(ln = c(0.1,0.9), rlab = NULL))
    colMeans(scores)
}))
```

The PIT histogram is based on the pointwise ECDF of the simulated epidemic
curves.

```{r hhh4sims_pit, fig.width = 6, fig.height = 3, echo = -1}
par(mar = c(5,5,1,1), mfrow = c(1,3))
invisible(lapply(hhh4sims, function (x) {
    pit(x = x$observed, pdistr = apply(x$sims, 1, ecdf))
}))
```
