---
title: "Forecasting Swiss ILI counts using simple log-normals by calendar week"
author: "Sebastian Meyer"
date: "`r Sys.Date()`"
output:
    rmarkdown::html_vignette:
        fig_width: 6
        fig_height: 4
        toc: TRUE
vignette: >
  %\VignetteIndexEntry{Forecasting Swiss ILI counts using simple log-normals by calendar week}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{ggplot2, fanplot, MASS}
---

```{r setup_knitr, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, error = FALSE,
                      fig.align = "center", dev.args = list(pointsize = 10))
```

```{r setup}
options(digits = 4)  # for more compact numerical outputs
library("HIDDA.forecasting")
library("ggplot2")
source("setup.R", local = TRUE)  # define test periods (OWA, TEST)
```

In this vignette, we compute naive reference forecasts to be compared with
the more sophisticated modelling approaches presented in the other vignettes.
Our naive approach to predict weekly ILI counts from
`r format_period(OWA)` (the `OWA` period) is to estimate a log-normal
distribution from the counts observed in the previous years in the same
calendar week. At each week, we estimate the two parameters using maximum
likelihood as implemented in **MASS**`::fitdistr()`.

The corresponding software reference is:
```{r, echo = FALSE, results = "asis"}
cat("<blockquote>")
print(citation(package = "MASS", auto = TRUE), style = "html")
cat("</blockquote>\n")
```


## One-week-ahead forecasts

```{r naiveowa}
CHILI_calendarweek <- as.integer(strftime(index(CHILI), "%V"))
naiveowa <- t(sapply(X = OWA+1, FUN = function (week) {
    cw <- CHILI_calendarweek[week]
    index_cws <- which(CHILI_calendarweek == cw)
    index_prior_cws <- index_cws[index_cws < week]
    MASS::fitdistr(CHILI[index_prior_cws], "lognormal")$estimate
}))
```

```{r naiveowa_pit, fig.width = 3, fig.height = 3, echo = -1}
par(mar = c(5,5,1,1), las = 1)
.PIT <- plnorm(CHILI[OWA+1], meanlog = naiveowa[,"meanlog"], sdlog = naiveowa[,"sdlog"])
hist(.PIT, breaks = seq(0, 1, 0.1), freq = FALSE, main = "", xlab = "PIT")
abline(h = 1, lty = 2, col = "grey")
```

```{r naiveowa_scores}
naiveowa_scores <- scores_lnorm(x = CHILI[OWA+1],
                                meanlog = naiveowa[,"meanlog"],
                                sdlog = naiveowa[,"sdlog"],
                                which = c("dss", "logs"))
summary(naiveowa_scores)
```

```{r naiveowa_plot, echo = -2}
naiveowa_quantiles <- sapply(X = 1:99/100, FUN = qlnorm,
                             meanlog = naiveowa[,"meanlog"],
                             sdlog = naiveowa[,"sdlog"])
par(mar = c(5,5,1,1))
osaplot(
    quantiles = naiveowa_quantiles, probs = 1:99/100,
    observed = CHILI[OWA+1], scores = naiveowa,
    start = OWA[1]+1, xlab = "Week", ylim = c(0,60000),
    fan.args = list(ln = c(0.1,0.9), rlab = NULL)
)
```


## Long-term forecasts

With this naive forecasting approach, the long-term forecast for a whole
season is simply composed of the sequential one-week-ahead forecasts
during that season.

```{r naivefor}
rownames(naiveowa) <- OWA+1
naivefor <- lapply(TEST, function (testperiod) {
    owas <- naiveowa[as.character(testperiod),,drop=FALSE]
    list(testperiod = testperiod,
         observed = as.vector(CHILI[testperiod]),
         meanlog = owas[,"meanlog"], sdlog = owas[,"sdlog"])
})
```

```{r naivefor_pit, echo = -1}
par(mar = c(5,5,1,1), mfrow = sort(n2mfrow(length(naivefor))), las = 1)
invisible(lapply(naivefor, function (x) {
    PIT <- plnorm(x$observed, meanlog = x$meanlog, sdlog = x$sdlog)
    hist(PIT, breaks = seq(0, 1, 0.1), freq = FALSE,
         main = format_period(x$testperiod, fmt = "%Y", collapse = "/"))
    abline(h = 1, lty = 2, col = "grey")
}))
```

```{r naivefor_plot, echo = -1, fig.show = "hold"}
par(mar = c(5,5,1,1))
t(sapply(naivefor, function (x) {
    quantiles <- sapply(X = 1:99/100, FUN = qlnorm,
                        meanlog = x$meanlog, sdlog = x$sdlog)
    scores <- scores_lnorm(x = x$observed,
                           meanlog = x$meanlog, sdlog = x$sdlog,
                           which = c("dss", "logs"))
    osaplot(quantiles = quantiles, probs = 1:99/100,
            observed = x$observed, scores = scores,
            start = x$testperiod[1], xlab = "Week", ylim = c(0,60000),
            fan.args = list(ln = c(0.1,0.9), rlab = NULL))
    colMeans(scores)
}))
```
