---
title: "Forecasting Swiss ILI surveillance counts using `forecast::auto.arima`"
author: "Sebastian Meyer"
date: "`r Sys.Date()`"
output:
    rmarkdown::html_vignette:
        fig_width: 7
        fig_height: 4
        toc: TRUE
vignette: >
  %\VignetteIndexEntry{Forecasting Swiss ILI surveillance counts using 'forecast::auto.arima'}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{ggplot2, surveillance, fanplot, forecast}
---

```{r setup_knitr, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, error = FALSE)
```

```{r setup}
library("HIDDA.forecasting")
library("ggplot2")
source("setup.R")  # define TRAIN, TEST and OWA periods
```


## Modelling

```{r}
library("forecast")
```

ARIMA models assume a Gaussian response, so we need to work with transformed counts:

```{r BoxCox.lambda}
BoxCox.lambda(CHILI, method = "loglik")
```

=> Box-Cox procedure suggests a log-transformation (`lambda = 0`).


```{r arimafit, cache = TRUE}
arimafit <- auto.arima(CHILI, lambda = 0, stepwise = FALSE, approximation = FALSE)
summary(arimafit)
```

CAVE: Seasonality cannot be taken into account in the standard way because the
data have no regular frequency (not a standard `"ts"`) ...
But we can add sine/cosine covariates just like in the endemic part of the above
`hhh4` model.

```{r sarimafit, cache = TRUE}
sarima_cov <- t(sapply(2*pi*seq_along(CHILI)/52.1775,
                       function (x) c(sin = sin(x), cos = cos(x))))
sarimafit <- auto.arima(CHILI, lambda = 0, stepwise = FALSE, approximation = FALSE,
                        xreg = sarima_cov)
summary(sarimafit)
```

```{r sarimaxfit, cache = TRUE}
sarimax_cov <- cbind(sarima_cov,
                     christmas = as.integer(strftime(index(CHILI), "%V") == "52"))
sarimaxfit <- auto.arima(CHILI, lambda = 0, stepwise = FALSE, approximation = FALSE,
                         xreg = sarimax_cov)
summary(sarimaxfit)
```


```{r}
CHILIdat <- fortify(CHILI)
CHILIdat$sarimaxfitted <- fitted(sarimaxfit)
CHILIdat <- cbind(CHILIdat,
    sapply(c(sarimaxlower=0.025, sarimaxupper=0.975), function (p)
        InvBoxCox(BoxCox(fitted(sarimaxfit), lambda = sarimaxfit$lambda) +
                  qnorm(p, sd = sqrt(sarimaxfit$sigma2)),
                  lambda = sarimaxfit$lambda)))
```

```{r sarimaxfitted}
ggplot(CHILIdat, aes(x=Index, ymin=sarimaxlower, y=sarimaxfitted, ymax=sarimaxupper)) +
    geom_ribbon(fill="orange") + geom_line(col="darkred") +
    geom_point(aes(y=CHILI), pch=20) +
    scale_y_sqrt(expand = c(0,0), limits = c(0,NA))
```



## One-week-ahead forecasts

The model selected above is refitted at each time point, but
we do not repeat `auto.arima()` model selection.
This is similar to so-called time-series cross-validation as implemented in
`forecast::tsCV()`. However, `tsCV()` only computes absolute errors of the point
forecasts, whereas we are interested in assessing probabilistic forecasts so
also need the forecast variance.

For each time point, forecasting with `Arima` takes about 0.5 seconds, i.e.,
computing `r length(OWA)` one-week-ahead forecasts takes approx.
`r sprintf("%.1f", length(OWA) * 0.5/60)` minutes ... but we can parallelize.

```{r, include = FALSE, eval = FALSE}
## check update.Arima: we obtain the same fit if we don't change the subset
all.equal(update(sarimaxfit)[names(sarimaxfit)], sarimaxfit)
```

```{r sarimaxowa, eval = !file.exists("sarimaxowa.RData"), results = "hide"}
sarimaxowa <- t(simplify2array(surveillance::plapply(X = OWA, FUN = function (t) {
    sarimaxfit_t <- update(sarimaxfit, subset = 1:t)
    unlist(predict(sarimaxfit_t, nahead=1, newxreg=sarimax_cov[t+1,,drop=FALSE]))
}, .parallel = 2)))
save(sarimaxowa, file = "sarimaxowa.RData")
```
```{r, include = FALSE}
load("sarimaxowa.RData")
```


ARIMA forecasts for the log-counts are normal with mean `pred` and variance `se^2`
=> back-transformation via exp() is log-normal

```{r sarimaxowa_pit}
surveillance::pit(
    x = CHILI[OWA+1], pdistr = plnorm,
    meanlog = sarimaxowa[,"pred"], sdlog = sarimaxowa[,"se"],
    plot = list(main = "PIT histogram of one-week-ahead forecasts from ARIMA")
)
```

```{r sarimaxowa_scores}
sarimaxowa_scores <- scores_lnorm(x = CHILI[OWA+1],
                                  meanlog = sarimaxowa[,"pred"],
                                  sdlog = sarimaxowa[,"se"],
                                  which = c("dss", "logs"))
summary(sarimaxowa_scores)
```

```{r sarimaxowa_plot}
sarimaxowa_quantiles <- sapply(X = 1:99/100, FUN = qlnorm,
                               meanlog = sarimaxowa[,"pred"],
                               sdlog = sarimaxowa[,"se"])
par(mar = c(5,5,1,1))
osaplot(
    quantiles = sarimaxowa_quantiles, probs = 1:99/100,
    observed = CHILI[OWA+1], scores = sarimaxowa_scores,
    start = OWA[1]+1, xlab = "Week", ylim = c(0,60000)
)
```



## Long-term forecasts

```{r}
## use the first forecast period for testing
TRAIN1 <- TRAIN[[1]]
TEST1 <- TEST[[1]]
```

```{r sarimaxfor}
fit1 <- update(sarimaxfit, subset = TRAIN1)
sarimaxfor <- predict(fit1, n.ahead = length(TEST1), newxreg = sarimax_cov[TEST1,,drop=FALSE])
```
```{r, include = FALSE}
## check that quantiles from predict() agree with quantiles from forecast()
q <- sapply(X = c(0.1,0.025,0.9,0.975), FUN = qlnorm,
            meanlog = sarimaxfor$pred, sdlog = sarimaxfor$se)
fc <- forecast(fit1, xreg = sarimax_cov[TEST1,,drop=FALSE], fan = FALSE)
stopifnot(all.equal(q, unclass(cbind(fc$lower, fc$upper)), check.attributes = FALSE))
```

```{r sarimaxfor_pit}
surveillance::pit(
    x = CHILI[TEST1], pdistr = plnorm,
    meanlog = sarimaxfor$pred, sdlog = sarimaxfor$se,
    plot = list(main = "PIT histogram of long-term forecast from ARIMA")
)
```

```{r sarimaxfor_scores}
sarimaxfor_scores <- scores_lnorm(x = CHILI[TEST1],
                                  meanlog = unclass(sarimaxfor$pred),
                                  sdlog = unclass(sarimaxfor$se),
                                  which = c("dss", "logs"))
summary(sarimaxfor_scores)
```

```{r sarimaxfor_plot}
sarimaxfor_quantiles <- sapply(X = 1:99/100, FUN = qlnorm,
                               meanlog = sarimaxfor$pred, sdlog = sarimaxfor$se)
par(mar = c(5,5,1,1))
osaplot(
    quantiles = sarimaxfor_quantiles, probs = 1:99/100,
    observed = CHILI[TEST1], scores = sarimaxfor_scores,
    start = TEST1[1], xlab = "Week", ylim = c(0,60000)
)
```
