---
title: "Forecasting Swiss ILI surveillance counts using `prophet::prophet`"
author: "Sebastian Meyer"
date: "`r Sys.Date()`"
output:
    rmarkdown::html_vignette:
        fig_width: 7
        fig_height: 4
        toc: TRUE
vignette: >
  %\VignetteIndexEntry{Forecasting Swiss ILI surveillance counts using 'prophet::prophet'}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{ggplot2, surveillance, fanplot, prophet}
---

```{r setup_knitr, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, error = FALSE)
```

```{r setup}
library("HIDDA.forecasting")
library("ggplot2")
source("setup.R")  # define TRAIN, TEST and OWA periods
```


## Modelling

```{r}
library("prophet")
```

```{r}
## which dates correspond to ISO week 52?
index(CHILI)[strftime(index(CHILI), "%V") == "52"]
```

So for consistency with the above models we need to define
holidays from 21 to 28 December each year

```{r prophetfit, cache = TRUE}
prophetfit_control <- prophet(
    yearly.seasonality = TRUE, weekly.seasonality = FALSE, daily.seasonality = FALSE,
    holidays = data.frame(
        holiday = "Christmas",
        ds = as.Date(paste0(2000:2016, "-12-21")),
        lower_window = 0,
        upper_window = 7
    ),
    mcmc.samples = 0, # invokes rstan::optimizing (fast MAP estimation)
    interval.width = 0.95, fit = FALSE)
prophetfit <- fit.prophet(
    m = prophetfit_control,
    df = data.frame(ds = index(CHILI), y = log(CHILI))
)
prophetfitted <- predict(prophetfit)
```

```{r prophetfitted_components}
prophet_plot_components(prophetfit, prophetfitted)
```

```{r}
CHILIdat <- fortify(CHILI)
CHILIdat[c("prophetfitted","prophetlower","prophetupper")] <-
    exp(prophetfitted[c("yhat", "yhat_lower", "yhat_upper")])
```

```{r prophetfitted}
##plot(prophetfit, prophetfitted)
ggplot(CHILIdat, aes(x=Index, ymin=prophetlower, y=prophetfitted, ymax=prophetupper)) +
    geom_ribbon(fill="orange") + geom_line(col="darkred") +
    geom_point(aes(y=CHILI), pch=20) +
    scale_y_sqrt(expand = c(0,0), limits = c(0,NA))
```


## One-week-ahead forecasts

For each time point, forecasting with `prophet` takes about 3 seconds, i.e.,
computing `r length(OWA)` one-week-ahead forecasts takes approx.
`r sprintf("%.1f", length(OWA) * 3/60)` minutes ...

```{r prophetowa, eval = !file.exists("prophetowa.RData"), results = "hide"}
prophetfitall <- fit.prophet(
    m = prophetfit_control,
    df = data.frame(ds = index(CHILI), y = log(CHILI))
)
prophetowa <- simulated_historical_forecasts(
    model = prophetfitall, k = length(OWA),
    period = 1, horizon = 1, units = "weeks"
)
## add forecast variance (calculated from prediction interval)
prophetowa$sigma <- with(prophetowa, (yhat_upper-yhat_lower)/2/qnorm(0.975))
## save results
save(prophetowa, file = "prophetowa.RData")
```
```{r, include = FALSE}
load("prophetowa.RData")
```

```{r, include = FALSE}
## BTW, the error SD can be obtained from a prophet fit as
prophetfit$y.scale * prophetfit$params$sigma_obs
```


`prophet` forecasts for the log-counts are approximately Gaussian with mean
`yhat` and variance `sigma^2` => back-transformation via exp() is log-normal

```{r prophetowa_pit}
surveillance::pit(
    x = exp(prophetowa$y), pdistr = plnorm,
    meanlog = prophetowa$yhat, sdlog = prophetowa$sigma,
    plot = list(main = "PIT histogram of one-week-ahead forecasts from prophet")
)
```

```{r prophetowa_scores}
prophetowa_scores <- scores_lnorm(x = exp(prophetowa$y),
                                  meanlog = prophetowa$yhat,
                                  sdlog = prophetowa$sigma,
                                  which = c("dss", "logs"))
summary(prophetowa_scores)
```

```{r prophetowa_plot}
prophetowa_quantiles <- sapply(X = 1:99/100, FUN = qlnorm,
                               meanlog = prophetowa$yhat,
                               sdlog = prophetowa$sigma)
par(mar = c(5,5,1,1))
osaplot(
    quantiles = prophetowa_quantiles, probs = 1:99/100,
    observed = exp(prophetowa$y), scores = prophetowa_scores,
    start = OWA[1]+1, xlab = "Week", ylim = c(0,60000)
)
```


## Long-term forecasts

```{r}
## use the first forecast period for testing
TRAIN1 <- TRAIN[[1]]
TEST1 <- TEST[[1]]
```

```{r}
##future <- make_future_dataframe(prophetfit, periods = length(TEST1), freq = "week", include_history = FALSE)
##forecast <- predict(prophetfit, data.frame(ds = index(CHILI)[TEST1]))
```
