---
title: "Long-term forecasts of age-stratified norovirus incidence in Berlin"
author: "Sebastian Meyer"
date: "`r Sys.Date()`"
output:
    rmarkdown::html_vignette:
        fig_width: 6
        fig_height: 4
        toc: TRUE
vignette: >
  %\VignetteIndexEntry{Long-term forecasts of age-stratified norovirus incidence in Berlin}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{surveillance, hhh4contacts, hhh4addon, fanplot}
---

```{r setup_knitr, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, error = FALSE,
                      fig.align = "center", dev.args = list(pointsize = 10))
options(knitr.duplicate.label = "allow")
```

```{r setup}
options(digits = 4)  # for more compact numerical outputs
library("surveillance")
library("hhh4contacts")
```

In this vignette, we assess long-term forecasts of the "hhh4" models from
`vignette("BNV")` using:
```{r}
library("hhh4addon")
```
This package is developed by Johannes Bracher and available from
`r packageDescription("hhh4addon", fields = "URL")`.


## Models

We rerun the code from the `vignette("BNV")` to make the models available:

```{r BNV, results = "hide", cache = TRUE}
rmarkdown::render(system.file("doc", "BNV.Rmd", package = "HIDDA.forecasting"),
                  run_pandoc = FALSE, envir = BNV <- new.env(), quiet = TRUE)
```

```{r}
(MODELS <- names(BNV$fits))
NMODELS <- length(MODELS)
```


## Predictive distributions

```{r stationary_distributions, eval = FALSE, include = FALSE}
statmoms <- lapply(BNV$fits, stationary_moments, start = 27)

## illustration for age group 00-04
.group <- 1
par(mfrow = n2mfrow(NMODELS), mar = c(5,5,2,1))
invisible(mapply(fanplot_stationary, statmoms, main = names(statmoms),
                 MoreArgs = list(unit = .group, xlab = "Time within season",
                                 mean_col = "white", mean_lty = 1)))
```

```{r predictive_distributions}
predmoms <- lapply(BNV$fits, predictive_moments, t_condition = 208, lgt = 52,
                   return_Sigma = TRUE)

## illustration for age group 00-04
.group <- 1
par(mfrow = n2mfrow(NMODELS), mar = c(5,5,2,1), las = 1)
invisible(mapply(fanplot_prediction, predmoms, main = names(predmoms),
                 MoreArgs = list(unit = .group, xlab = "Time", ylim = c(0, 55),
                                 mean_col = "white", mean_lty = 1,
                                 l.col = NA, pt.col = 1, pt.cex = 0.4)))
```

Scaled Dawid-Sebastiani scores of these multivariate forecast distributions:

```{r DSS}
sapply(predmoms, ds_score_hhh4)
```


### Final size forecasts (aggregate over time)

```{r predictive_distributions_aggregated}
aggr <- matrix(rep(diag(BNV$NGROUPS), 52), nrow = BNV$NGROUPS,
               dimnames = list(BNV$GROUPS, NULL))
predmoms_aggr <- lapply(predmoms, aggregate_moments, aggregation_matrix = aggr)

par(mfrow = n2mfrow(NMODELS), mar = c(5,5,2,1), las = 2, cex.axis = 0.8)
invisible(mapply(plot_moments_by_unit, predmoms_aggr, main = names(predmoms_aggr),
                 MoreArgs = list(xlab = "Age group", ylim = c(0, 3000),
                                 pt.col = 2, pt.cex = 0.6)))
```

Scaled Dawid-Sebastiani scores of these multivariate forecast distributions:

```{r include = FALSE}
## workaround for a bug in "hhh4addon"
predmoms_aggr <- mapply("[[<-", predmoms_aggr, "mu_vector",
                        lapply(predmoms_aggr, "[[", "mu"), SIMPLIFY = FALSE)
```

```{r DSS_aggr}
sapply(predmoms_aggr, ds_score_hhh4)
```
