---
title: "Forecasting Swiss ILI surveillance counts using `tscount::tsglm`"
author: "Sebastian Meyer"
date: "`r Sys.Date()`"
output:
    rmarkdown::html_vignette:
        fig_width: 6
        fig_height: 4
        toc: TRUE
vignette: >
  %\VignetteIndexEntry{Forecasting Swiss ILI surveillance counts using 'tscount::tsglm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{ggplot2, surveillance, fanplot, tscount}
---

```{r setup_knitr, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, error = FALSE,
                      fig.align = "center", dev.args = list(pointsize = 10))
```

```{r setup}
library("HIDDA.forecasting")
library("ggplot2")
source("../setup.R", local = TRUE)  # define test periods (OWA, TEST)
```

In this vignette, we use forecasting methods provided by:
```{r}
library("tscount")
```

The corresponding software reference is:
```{r, echo = FALSE, results = "asis"}
cat("<blockquote>")
print(citation(package = "tscount", auto = TRUE), style = "html")
cat("</blockquote>\n")
```


## Modelling

Construct the matrix of covariates, including yearly seasonality and a Christmas
effect as for the other models (see, e.g., `vignette("CHILI_hhh4")`):

```{r xreg}
X <- t(sapply(2*pi*seq_along(CHILI)/52.1775,
              function (x) c(sin = sin(x), cos = cos(x))))
X <- cbind(X,
           christmas = as.integer(strftime(index(CHILI), "%V") == "52"))
```

Fitting a NegBin "time-series GLM" with a log-link, regressing on the
counts from the last three weeks (autoregression) and on the conditional
mean 52 weeks ago (for seasonality, so we omit the sine-cosine covariates):

```{r tsglmfit, fig.height = 6}
tsglmfit <- tsglm(CHILI, model=list(past_obs=1:3, past_mean=52),
                  xreg=X[,"christmas",drop=FALSE], distr="nbinom", link="log")
summary(tsglmfit)
```

```{r}
CHILIdat <- fortify(CHILI)
CHILIdat$tsglmfitted <- fitted(tsglmfit)
CHILIdat <- cbind(CHILIdat,
    sapply(c(tsglmlower=0.025, tsglmupper=0.975), function (p)
        qnbinom(p, mu = fitted(tsglmfit), size = tsglmfit$distrcoefs)))
```

```{r tsglmfitted, fig.width = 7, fig.height = 4}
ggplot(CHILIdat, aes(x=Index, ymin=tsglmlower, y=tsglmfitted, ymax=tsglmupper)) +
    geom_ribbon(fill="orange") + geom_line(col="darkred") +
    geom_point(aes(y=CHILI), pch=20) +
    scale_y_sqrt(expand = c(0,0), limits = c(0,NA))
```



## One-week-ahead forecasts


We compute `r length(OWA)` one-week-ahead forecasts
from `r format_period(OWA)` (the `OWA` period).
The model is refitted at each time point.

For each time point, refitting and forecasting with `tsglm` takes about 3 seconds, i.e.,
computing all one-week-ahead forecasts takes approx.
`r sprintf("%.1f", length(OWA) * 3/60)` minutes ... but we can parallelize.

```{r tsglmowa, eval = !file.exists("tsglmowa.RData"), results = "hide"}
tsglmowa <- t(simplify2array(surveillance::plapply(X = OWA, FUN = function (t) {
    tsglmfit_t <- update(tsglmfit, ts = tsglmfit$ts[1:t],
                         xreg = tsglmfit$xreg[1:t,,drop=FALSE])
    c(mu = predict(tsglmfit_t, n.ahead = 1, newxreg = tsglmfit$xreg[t+1,,drop=FALSE])$pred,
      size = tsglmfit_t$distrcoefs[[1]])
}, .parallel = 3)))
save(tsglmowa, file = "tsglmowa.RData")
```
```{r, include = FALSE}
load("tsglmowa.RData")
```

```{r tsglmowa_pit, fig.width = 3, fig.height = 3, echo = -4}
## CAVE: tscount's pit() only uses the first of 'distrcoefs'
## pit(response = CHILI[OWA+1], pred = tsglmowa[,"mu"],
##     distr = "nbinom", distrcoefs = tsglmowa[,"size"])
par(mar = c(5,5,1,1), las = 1)
surveillance::pit(
    x = CHILI[OWA+1], pdistr = pnbinom,
    mu = tsglmowa[,"mu"], size = tsglmowa[,"size"],
)
```

```{r tsglmowa_scores}
## CAVE: tscount's scoring() only uses the first of 'distrcoefs'
## scoring(response = CHILI[OWA+1], pred = tsglmowa[,"mu"],
##         distr = "nbinom", distrcoefs = tsglmowa[,"size"])
tsglmowa_scores <- surveillance::scores(
    x = CHILI[OWA+1], mu = tsglmowa[,"mu"],
    size = tsglmowa[,"size"], which = c("dss", "logs"))
summary(tsglmowa_scores)
```

```{r tsglmowa_plot, echo = -2}
tsglmowa_quantiles <- sapply(X = 1:99/100, FUN = qnbinom,
                              mu = tsglmowa[,"mu"],
                              size = tsglmowa[,"size"])
par(mar = c(5,5,1,1))
osaplot(
    quantiles = tsglmowa_quantiles, probs = 1:99/100,
    observed = CHILI[OWA+1], scores = tsglmowa_scores,
    start = OWA[1]+1, xlab = "Week", ylim = c(0,60000),
    fan.args = list(ln = c(0.1,0.9), rlab = NULL)
)
```
